The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2015-01-22 draft.tex
approaching final draft

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles associated with ultraviolet
(UV) resonance lines, such as C~{\sc iv}~$\lambda1550$. These features
imply the existence of powerful accretion disk winds in
CVs.
% and considerable effort has gone into modelling UV line formation
% in such outflows. By contrast, much less attention has been paid to
% the potential impact of these disk winds on the {\em optical} spectra
% of CVs. 
Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the {\em optical} waveband. 
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs also leaves a noticeable imprint on
the optical spectrum, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombination lines are \ha\ and
\heiiopt, although the outflow also produces emission in the Ly$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. We demonstrate that a higher-density outflow model 
displays reasonable verisimilitude with the optical spectrum of RW Trianguli,
producing all the observed lines. This model
produces recombination continuum sufficient to fill in the Balmer jump,
but it also predicts stronger-than-observed 
collisionally excited UV emission lines. 
The wind in our revised model also alters the spectral shape of the continuum, 
and displays a strong inclination dependence in its emission properties. 
No single-peaked emission is seen, although we observe a 
narrowing of the double-peaked emission lines from the base of the wind.
We show that denser models can produce a single-peaked \ha\ line.
The single-peaked line profile is due to the line emission originating from 
$\sim10^11$~cm above the disk plane where rotational velocities are small, 
and is not due to the radial velocity shear mechanism that
has previously been proposed. 
On the basis of our results, we suggest that winds can 
modify, and perhaps even dominate, the line and continuum emission
from CVs.
% We carry out a limited exploration of
% the disk wind parameter space, focusing on those parameters that affect the
% density near the base of the wind. This shows that higher-density
% outflows can produce both single-peaked emission lines and a
% recombination continuum sufficient to fill in the Balmer
% jump, although such models also tend to predict stronger-than-observed 
% collisionally excited UV emission lines. Our main conclusion is that
% disk wind emission is likely to have a significant impact on the
% optical spectra of high-state CVs. 
% {\bf JM: This abstract is a few lines longer than I would like- 
% I think something punchier is required.}
\end{abstract}

@@

Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) 
has shown that it is possible to calculate the emissivity of a gas in
statistical equilibrium accurately by quantising matter into
`macro-atoms', and radiant and kinetic energy into indivisible energy
packets (r- and k- packets, respectively). His macro-atom scheme
allows for all possible transition paths from a given level and
provides a full non-local thermodynamic equilibrium (NLTE) solution
for the level populations based on Monte Carlo estimators. The macro-atom
technique has already been used to model Wolf-Rayet star
winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kromersim, kerzendorfsim} and YSOs (SDL05). A full 
description of the approach can be found in L02 and L03.

@@

In our implementation of the macro-atom approach, we also explicitly
take into account the photoionization and collisional ionization rates
between a lower level, $l$, and the continuum (or, in the case of ions
with more than one bound electron, the ground state of the upper ion),
$\kappa$,
\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu}
  \sigma_{\nu}}{h \nu} d\nu + C_{l \kappa} n_l n_e.
\end{equation}
Here, $\sigma_{\nu}$ is the photoionization cross section, and $J_{\nu}$
is the mean intensity. The corresponding recombination rate is given
by 
\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e + C_{\kappa l}
n_\kappa n_e, \\
\end{equation}
where $\alpha_{\kappa l}$ is the radiative recombination coefficient
to level $l$. This treatment means that radiative and collisional
rates to and from all levels are considered when calculating both the
ionization state and the level populations, although we neglect 
ionization directly to excited levels of the upper ion. The
Van Regemorter (1962) approximation is used for collisional
transitions, and thus we assume that collisions between radiatively
forbidden transitions are not important in determining level
populations. This is a reasonable approximation in the regimes
considered here.

@@

The density at position $(r,z)$ in the wind, $\rho(r,z)$, is
calculated from the mass continuity equation, yielding
\begin{equation}
\rho(r,z) = \frac{r_0}{r} \frac{dr_0}{dr} \frac{\phi(r_0)}{v_z(r,z)}.
\label{density}
\end{equation}
Here, 
$v_z$ is the vertical velocity component and $\phi(r_0)$ 
is the local mass-loss rate per unit area at $r_0$, 
%in the direction of a streamline (as defined by SV93), 
such that 
\begin{equation}
\phi(r_0) \propto \dot{M}_{wind} R_0^\lambda \cos [\theta(r_0)].
\label{density}
\end{equation}
We adopt $\lambda = 0$ and normalize $\phi(r_0)$ by 
matching its integral over both sides of the disk
to the user-specified total mass-loss rate, $\dot{M}_{wind}$. 
% {\bf CK: I'm still not sure I understand the formula
%   SV93 use to define $\phi$, and hence that this description is
%   actually strictly correct.... Not sure what to do about that
%   though...}

@@

The general behaviour we might expect is that outflows with denser
regions near the wind base - i.e. winds with larger $R_{v}$ and/or
larger $\alpha$ - will produce stronger optical emission signatures. 
However, this behaviour may be moderated by the effect of the increasing
optical depth through this region, which can also affect the line profile shapes. 
In addition, modifyin $R_v$ in particular also tends to increase the emission volume.
Having briefly explored a variety of models with different kinematics,
we adopt the parameters listed in table~\ref{modelb_table}
for our revised model (model B), which is optimized for the optical waveband.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} emission line at 3934\AA\ 
becomes quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the He
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~$\lambda4686$ line (the He
equivalent of Paschen~$\alpha$). 
%%Although it has been observed (REF),
This line is somewhat unfamiliar observationally, because it 
lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

Our models do not exhibit P-Cygni profiles in the optical lines.
This is perhaps not surprising. LK02 and SV93 originally designed such models
to reproduce the UV line profiles. Thus, most of the wind
has an ionization parameter of $\log U \sim 2$ (see Fig.~\ref{wind}).
This means H and He are fully ionized throughout 
much of the wind and are successful in producing recombination features.
However, the line opacity throughout the wind is too
low to produce noticeable blue shifted absorption. 
We suspect that the winds that exhibit such profiles must 
possess a higher degree of ionization stratification, although the lack 
of contemporary observations means it is not known for certain if the 
P-Cygni profiles in UV resonance lines and optical H and He lines exist simultaneously.
Ionization stratification could manifest as a clumpy flow, in which ionization changes
due to small scale density fluctuations, or a stratification in density
and ionizing radiation field over larger scales.
Theory of line-driven winds expects them to be unsteady (REFs)
and modelling of CV winds also predicts density inhomogeneities (REF).
Tentative evidence for clumping being directly related to P-Cygni optical lines
comes from the fact that \cite{prinja2000}
found dwarf nova BZ Cam's outflow to be unsteady and highly mass-loaded in outburst,
by observing the UV resonance lines.
This system has exhibited P-Cygni profiles in He~\textsc{i}~$\lambda5876$
and \ha\ when in a high-state \citep{patterson1996,RN98}. 
We also note that the degree of ionization and density variation and 
subsequent line opacities may be affected by our model parameters
and the specific parameterisation adopted.

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in velocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_v = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a narrowing of the lines, and a single-peaked line in model X.
This is not due to radial velocity shear (see section 5.3).
}
\label{halpha}
\end{figure} %fullpage

>>>>

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles associated with ultraviolet
(UV) resonance lines, such as C~{\sc iv}~$\lambda1550$. These features
imply the existence of powerful accretion disk winds in
CVs.
% and considerable effort has gone into modelling UV line formation
% in such outflows. By contrast, much less attention has been paid to
% the potential impact of these disk winds on the {\em optical} spectra
% of CVs. 
Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the {\em optical} waveband. 
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs also leaves a noticeable imprint on
the optical spectrum, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombination lines are \ha\ and
\heiiopt, although the outflow also produces emission in the Ly$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. We demonstrate that a higher-density outflow model 
displays reasonable verisimilitude with the optical spectrum of RW Trianguli,
producing all the observed lines. This model
produces recombination continuum sufficient to fill in the Balmer jump,
but it also predicts stronger-than-observed 
collisionally excited UV emission lines. 
The wind in our revised model also alters the spectral shape of the continuum, 
and displays a strong inclination dependence in its emission properties. 
No single-peaked emission is seen, although we observe a 
narrowing of the double-peaked emission lines from the base of the wind.
We show that denser models can produce a single-peaked \ha\ line.
The single-peaked line profile is due to the line emission originating from 
$\sim10^{11}$~cm above the disk plane where rotational velocities are small, 
and is not due to the radial velocity shear mechanism that
has previously been proposed. 
On the basis of our results, we suggest that winds can 
modify, and perhaps even dominate, the line and continuum emission
from CVs.
% We carry out a limited exploration of
% the disk wind parameter space, focusing on those parameters that affect the
% density near the base of the wind. This shows that higher-density
% outflows can produce both single-peaked emission lines and a
% recombination continuum sufficient to fill in the Balmer
% jump, although such models also tend to predict stronger-than-observed 
% collisionally excited UV emission lines. Our main conclusion is that
% disk wind emission is likely to have a significant impact on the
% optical spectra of high-state CVs. 
% {\bf JM: This abstract is a few lines longer than I would like- 
% I think something punchier is required.}
\end{abstract}

@@

Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) 
has shown that it is possible to calculate the emissivity of a gas in
statistical equilibrium accurately by quantising matter into
`macro-atoms', and radiant and kinetic energy into indivisible energy
packets (r- and k- packets, respectively). His macro-atom scheme
allows for all possible transition paths from a given level and
provides a full non-local thermodynamic equilibrium (NLTE) solution
for the level populations based on Monte Carlo estimators. The macro-atom
technique has already been used to model Wolf-Rayet star
winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kromersim2009, kerzendorfsim} and YSOs (SDL05). A full 
description of the approach can be found in L02 and L03.

@@

In our implementation of the macro-atom approach, we also explicitly
take into account the photoionization and collisional ionization rates
between a lower level, $l$, and the continuum (or, in the case of ions
with more than one bound electron, the ground state of the upper ion),
$\kappa$,
\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu}
  \sigma_{\nu}}{h \nu} d\nu + C_{l \kappa} n_l n_e.
\end{equation}
Here, $\sigma_{\nu}$ is the photoionization cross section, and $J_{\nu}$
is the mean intensity. The corresponding recombination rate is given
by 
\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e + C_{\kappa l}
n_\kappa n_e, \\
\end{equation}
where $\alpha_{\kappa l}$ is the radiative recombination coefficient
to level $l$. This treatment means that radiative and collisional
rates to and from all levels are considered when calculating both the
ionization state and the level populations, although we neglect 
ionization directly to excited levels of the upper ion. The
\cite{vanregemorter} approximation is used for collisional
transitions, and thus we assume that collisions between radiatively
forbidden transitions are not important in determining level
populations. This is a reasonable approximation in the regimes
considered here.

@@

The density at position $(r,z)$ in the wind, $\rho(r,z)$, is
calculated from the mass continuity equation, yielding
\begin{equation}
\rho(r,z) = \frac{r_0}{r} \frac{dr_0}{dr} \frac{\phi(r_0)}{v_z(r,z)}.
\label{density}
\end{equation}
Here, 
$v_z$ is the vertical velocity component and, following SV93,
$\phi(r_0)$ is the local mass-loss rate per unit area at $r_0$, 
such that 
\begin{equation}
\phi(r_0) \propto \dot{M}_{wind} r_0^\lambda \cos [\theta(r_0)].
\label{density}
\end{equation}
We adopt $\lambda = 0$ and normalize $\phi(r_0)$ by 
matching its integral over both sides of the disk
to the user-specified total mass-loss rate, $\dot{M}_{wind}$. 
% {\bf CK: I'm still not sure I understand the formula
%   SV93 use to define $\phi$, and hence that this description is
%   actually strictly correct.... Not sure what to do about that
%   though...}

@@

The general behaviour we might expect is that outflows with denser
regions near the wind base - i.e. winds with larger $R_{v}$ and/or
larger $\alpha$ - will produce stronger optical emission signatures. 
However, this behaviour may be moderated by the effect of the increasing
optical depth through this region, which can also affect the line profile shapes. 
In addition, modifying $R_v$ also increases the emission {\em volume}.
Having briefly explored a variety of models with different kinematics,
we adopt the parameters listed in table~\ref{modelb_table}
for our revised model (model B), which is optimized for the optical waveband.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} emission line at 3934\AA\ 
becomes quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the He
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~$\lambda4686$ line (the He
equivalent of Paschen~$\alpha$). 
This line is somewhat unfamiliar observationally, because it 
lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra.

Our models do not exhibit P-Cygni profiles in the optical lines.
This is perhaps not surprising. LK02 and SV93 originally designed such models
to reproduce the UV line profiles. Thus, most of the wind
has an ionization parameter of $\log U \sim 2$ (see Fig.~\ref{wind}).
This means H and He are fully ionized throughout 
much of the wind and are successful in producing recombination features.
However, the line opacity throughout the wind is too
low to produce noticeable blue shifted absorption. 
We suspect that the winds that exhibit such profiles must 
possess a higher degree of ionization stratification, although the lack 
of contemporary observations means it is not known for certain if the 
P-Cygni profiles in UV resonance lines and optical H and He lines exist simultaneously.
Ionization stratification could manifest as a clumpy flow, in which ionization changes
due to small scale density fluctuations, or a stratification in density
and ionizing radiation field over larger scales.
Theory of line-driven winds expects them to be unsteady \citep{macgregor1979,owockirybicki1984,owockirybicki1985},
simulations of CV disk winds also predict density inhomogeneities 
\citep{proga1998,pkdh2002}.
Tentative evidence for clumping being directly related to P-Cygni optical lines
comes from the fact that \cite{prinja2000}
found dwarf nova BZ Cam's outflow to be unsteady and highly mass-loaded in outburst,
by observing the UV resonance lines.
This system has exhibited P-Cygni profiles in He~\textsc{i}~$\lambda5876$
and \ha\ when in a high-state \citep{patterson1996,RN98}. 
We also note that the degree of ionization and density variation and 
subsequent line opacities may be affected by our model parameters
and the specific parameterisation adopted.

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalized to 1, plotted in velocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_v = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a narrowing of the lines, and a single-peaked line in model X.
This is not due to radial velocity shear (see section 5.3).
}
\label{halpha}
\end{figure} %fullpage
